---
output:
  xaringan::moon_reader:
    css: ["default", "styling/sydney-fonts.css", "styling/sydney.css"]
    self_contained: false
    seal: true
    nature:
      beforeInit: ["styling/remark-zoom.js"]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
      navigation:
        scroll: false
---

class: title-slide
background-image: url("styling/USydLogo-black.svg"), url("styling/title-image1.jpg")
background-position: 10% 90%, 100% 50%
background-size: 160px, 100% 100%

.content-box-purple[
# .black[Activity 1]
## Identifying Biomarkers and Making Predictions of Treatment Resistance
### Ellis Patrick
### 29 June 2018
]

---

## Activity Overview

- RNA-seq Data Set: Acute Myeloid Leukaemia Treatment Resistance

--

- ClassifyR: Biomarker Selection Methods and Cross-validation

--

- Differential Means and Differential Distribution Classifiers for Classifying AML Patients

---

## RNA-seq Data Set: Acute Myeloid Leukaemia (AML) Treatment Resistance

- Primary therapy resistance is a major problem in acute myeloid leukemia treatment. Approximately 20–30% of younger adult patients with AML and as many as 50% of older adults are refractory to induction treatment.

--

- Research findings are <a href="http://www.haematologica.org/content/103/3/456" target="_blank">published</a> in *Haematologica* in 2018.

--

- The data is available from <a href="https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE106291" target="_blank">GEO Browser</a> as 250 `txt.gz` files or a Microsoft Excel file where the gene expression values were standardised to have a mean of 0 and variance of 1.

--

```{r, echo = FALSE}
library(knitr)
opts_knit[["set"]](root.dir = "/home/dario/Documents/tutorial/")
```

Import the prepared tab-separated text files of RNA-seq read counts and clinical data.

```{r}
readCounts <- read.delim("data/counts.txt", check.names = FALSE)
readCounts <- as.matrix(readCounts)
sampleInfo <- read.delim("data/samples.txt", check.names = FALSE)
```

---

## Clinical Data

The cinical data provides information about seven different characteristics of the patients.

```{r}
# Add scroll box.
head(sampleInfo)
nrow(sampleInfo)
```

Observe the number of samples which are resistant to treatment and which are not.

```{r}
table(sampleInfo[, "Response"])
```

$164 + 71 = 235$, so there are 15 patients with missing data regarding their resistance. Identify which rows of the clinical data they are in.

```{r}
removeClinical <- which(is.na(sampleInfo[, "Response"]))
```

---

## AML Read Counts

```{r}
dim(readCounts)
```

There are `r nrow(readCounts)` genes in the counts table and `r ncol(readCounts)` AML patients.

--

```{r}
readCounts[1:6, 1:6]
```

---

## AML Read Counts

The number of reads per sample varies widely betwen patients by as much as 10-fold.

```{r}
samplesCounts <- colSums(readCounts)
countsSummary <- summary(samplesCounts)
countsSummary
```

--

Identify in which columns of the matrix patients with less than the first quartile of counted reads and more than the third quartile are to remove them from the analysis.

```{r}
removeRNA <- which(samplesCounts < countsSummary["1st Qu."] |
                   samplesCounts > countsSummary["3rd Qu."])
```

---

## Subsetting the Data Set

Remove samples which have a number of read counts below the first quartile of counted reads or above the third quartile as well as those missing resistance status.

```{r}
allRemove <- union(removeClinical, removeRNA)
readCounts <- readCounts[, -allRemove]
sampleInfo <- sampleInfo[-allRemove, ]
```

--

Caclulate the number of remanining samples belonging to each class.

```{r}
classes <- sampleInfo[, "Response"]
table(classes)
```

--

There is a moderate amount of class imbalance in this data set.

---

## Relationship of Mean and Variance

- RNA-seq count measurements are known to have a relationship between mean and variance.

- Many classification algorithms require data with constant variance over the range of means.

Observe the mean-variance relationship after scaling counts for different total counts per sample.

```{r, message = FALSE}
library(EDASeq)
AMLExpressionSet <- newSeqExpressionSet(readCounts)
AMLExpressionSet <- betweenLaneNormalization(AMLExpressionSet)
#meanVarPlot(AMLExpressionSet, log = TRUE, ylim = c(0, 20),
 #           main = "Mean-Variance Plot") bug; doesn't work on normalised counts!
```

---

Using DESeq2's `varianceStabilizingTransformation` function, remove the relationship of variance to mean.

```{r}
library(DESeq2)
measurementsVS <- varianceStabilizingTransformation(readCounts)
```

--

Visualise again the relationship between mean and variance.

```{r}
meanVarPlot(newSeqExpressionSet(measurementsVS), log = TRUE,
            xlim = c(1, 3), ylim = c(0, 3), main = "Mean-Variance Plot")
```

The variance is now constant across the range of means.

## Subsetting the Data Set Again

Before doing any classification, the number of genes will be reduced to make feature selection faster. The top 2000 most variable genes are retained.

```{r}
geneVariances <- apply(measurementsVS, 1, var)
mostVariable <- order(geneVariances, decreasing = TRUE)[1:2000]
measurementsVS <- measurementsVS[mostVariable, ]
head(measurementsVS)
```

Class information is not used, so the filtering is fair.

---

## ClassifyR

Introduce it.